---
title: "tidyr-notes"
author: "Tim Dennis"
date: "January 20, 2016"
output: html_document
---

Researchers want to manipulate their data from wide to long & visa-versa. 

 - each column is a variable
 - each row is an observation
 
* Long format you have 1 column for the observed vars and the other columns are ID vars
* Wide format each row is a site/subject/patient & multople observation varialbes containing the same type of data.
  * repeated observations over time
  * observations of multiple variables
* Data input might be easier in wide format (human readable)
* R's functions are designed assuming you have data in long format. 

```{r}
#install.packages('tidyr')
library(dplyr)
library(tidyr)
#str(gapminder)
```

* gapminder dataset has multiple observed vars 
* it is in between pure long and wide data format
  * 3 "ID Variables" (`continent`, `country`, `year`) 
  * 3 "observation varialbes" (`pop`, `lifeExp`, `gdpPercap`)
* dispite not having all observation vars in one column, this is ofte preferred b/c the observation vars are in different units.

## wide to long format with gather()

let's load the gap_wide data 

```{r}
gap_wide <- read.csv('https://goo.gl/4xPTex', header=TRUE, stringsAsFactors = FALSE)
str(gap_wide)
gapminder <- read.csv("https://goo.gl/BtBnPg", header = T)

```
* use gather to gather observations vars into a single variable


```{r}
gap_long <- gap_wide %>% 
  gather(obstype_year,obs_values,starts_with('pop'),starts_with('lifeExp'),starts_with('gdpPercap'))

str(gap_long)

```

* notice the use of the %>%  syntax as we used in dplyr. 
* yes you can combine tidyr and dplyr together using %>% 
* in gather() we name a new col for the new ID var (obstype_year) & the name for the amalgamated observation vars
* we use 'starts_with' to select all vars that start with that prefix
* gather also allows a `-` to indentify what vars are not to be gathered (ie ID vars)
* very useful for datasets with lots of vars

```{r}
gap_long <- gap_wide %>% gather(obstype_year, obs_values, -continent, -country)
str(gap_long)
```

* obstype_year actually contains 2 pieces of information, the observation type (pop,lifeExp, or gdpPercap) and the year
* use separate()

```{r}
gap_long <- gap_long %>% separate(obstype_year, into=c('obs_type', 'year'), sep='_')
gap_long$year <- as.integer(gap_long$year)
head(gap_long)
tail(gap_long)
```

## Long to intermediate

* use opposite of gather() to spread our data out again into intermediate width. 
* we can then go back to the widest format (what we started with above)

```{r}
gap_normal <- gap_long %>% spread(obs_type, obs_values)
dim(gap_normal)
```
```{r}
dim(gapminder)
names(gap_normal)
names(gapminder)
```
* Now we’ve got an intermediate dataframe gap_normal with the same dimensions as the original gapminder, but the order of the variables is different. 
* Let’s fix that before checking if they are all.equal().

 
```{r}
gap_normal <- gap_normal[, names(gapminder)]
all.equal(gap_normal, gapminder)
head(gap_normal)
head(gapminder)
```

```{r}
gap_normal <- gap_normal %>% arrange(country,continent,year)
all.equal(gap_normal,gapminder)
```

